### 2024-09-11 cognitive-bias.qmd
20240911 - narrative fallacy

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called *heuristics*.
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
For instance, snap judgments can help us rapidly differentiate between a friend and a foe, for survival.
However, these mental shortcuts can also lead us astray and to make systematic errors in our judgments and predictions.
*Cognitive biases* are systematic errors in thinking.
*Fallacies* are forms of flawed reasoning.

@@

*Hindsight bias* is the tendency to perceive that past events were more predictable than they were.
People tend to remember the succeess of their predictions and forget the failures of their predictions.
For instance, if a third-string Quarterback has a breakout game, a fantasy manager may claim that they "knew it all along" that the player was going to breakout, despite not having picked up the player.
That same manager may forget the many other predictions they had that did not come true.

@@

*Outcome bias* is the tendency to evaluate the quality of a decision based on the eventual outcome of that decision, rather than based on the quality of the decision based on the information that was known at the time of the decision.
For instance, consider that a manager starts a lower-ranked player over a higher-ranked player because the lower-ranked player has a more favorable matchup.
Now, consider that the lower-ranked player scores fewer points than the higher-ranked player because the lower-ranked player got injured.
If the manager concludes that their decision-making process for who to start was flawed and thus not pay attention to matchups in subsequent decisions, this would reflect outcome bias.
Likewise, if the manager makes a poor-quality decision but gets lucky, and they attribute their success to a sound decision-making process, this would also reflect outcome bias.

@@

>>>>

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called *heuristics*.
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
For instance, snap judgments can help us rapidly differentiate between a friend and a foe, for survival.
However, these mental shortcuts can also lead us astray and to make systematic errors in our judgments and predictions.
For instance, heuristics can lead us to identify patterns out of randomness.
*Cognitive biases* are systematic errors in thinking.
*Fallacies* are forms of flawed reasoning.

@@

*Hindsight bias* is the tendency to perceive that past events were more predictable than they were (i.e., the "I-knew-it-all-along" effect).
People tend to remember the succeess of their predictions and forget the failures of their predictions.
For instance, if a third-string Quarterback has a breakout game, a fantasy manager may claim that they "knew it all along" that the player was going to breakout, despite not having picked up the player.
That same manager may forget the many other predictions they had that did not come true.

@@

*Outcome bias* is the tendency to evaluate the quality of a decision based on the eventual outcome of that decision, rather than based on the quality of the decision based on the information that was known at the time of the decision.
For instance, consider that a manager starts a lower-ranked player over a higher-ranked player because the lower-ranked player has a more favorable matchup.
Now, consider that the lower-ranked player scores fewer points than the higher-ranked player because the lower-ranked player got injured.
If the manager concludes that their decision-making process for who to start was flawed and thus not pay attention to matchups in subsequent decisions, this would reflect outcome bias.
Likewise, if the manager makes a poor-quality decision but gets lucky, and they attribute their success to a sound decision-making process, this would also reflect outcome bias.
In general, people tend to blame decision makers for good decisions that turned out badly, and to give them too little credit for good decisions that turned out well [because the decisions appear obvious but only after the fact; @Kahneman2011].
[Hindsight bias](#sec-cognitiveBiasesHindsight) and outcome bias tend to lead to risk aversion [@Kahneman2011].

@@

### Narrative Fallacy {#sec-fallaciesNarrative}

The *narrative fallacy* is the error of creating a story with cause-and-effect explanations from random events.
According to @Kahneman2011, people find stories more compelling if they a) are simple (as opposed to complex), b) are concrete (as opposed to abstract), c) provide a greater attribution to talent, stupidity, and intentions (compared to luck), and d) focus on a few salient events that occured (rather than the many events that failed to occur).
For instance, consider that a player undergoes a challenging start to the season; however, they have higher-performing games in the middle of the season.
A fantasy football manager may construct a "causal narrative" that the player overcame obstacles and is now trending toward strong performance for the rest of the season, even though the stronger performance could reflect some natural and random fluctuation.
